---
title: "Week 2 Functions (basis of R files)"
author: '500455892'
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Packages

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(mvtnorm)
library(dr)
set.seed(2023)
source("sir_functions.R")
```

# (Week 2) SIR Functions

## Sorter

```{r}
sir_sorter_w2 <- function(dataset, y = colnames(dataset)[1]) {
  
  # This function sorts the dataset by the target variable
  
  # dataset is a dataframe
  # y is a character, that is the name of the column that is y (that you want to sort by)
  
  dataset_new <- dataset %>% arrange(get(y)) # arrange by values of y
  
  return(dataset_new) # return the dataset that is now sorted by values of y
}
```

## Divider

This takes a dataframe (that includes y) and a name for the column of y

```{r}
divider_w2 <- function(dataset, slices = 8, categorical = FALSE, y, ordinal = FALSE) {
  
  #### IMPORTANT: if y is categorical, must give arguments for dataset and y
  
  # Function to assign a slice number to each row of the dataset
  
  # Arguments: rows is an integer, should equal nrow(dataset);
  # slices is an integer, is the number of slices desired by user. Default is 8.
  # categorical is a boolean, value of TRUE set by user if y (from dataset) is a categorical variable, otherwise FALSE. If TRUE, then number of slices = number of unique categories. 
  
  # Method: create a vector whose length = rows, and assign a slice number to each element corresponding to which slice each row should belong to.
  
  rows <- nrow(dataset)
  
  if (!categorical) {
    rem <- rows%%slices # explain slices and set default 
    sizes <- rep(0, slices) # explain 
    for (i in 1:slices) {
      if (i <= rem) {
        sizes[i] <- rows%/%slices + 1
      } else {
        sizes[i] <- rows%/%slices
      }
    }
  } else {
    cols <- colnames(dataset)
    y_pos <- match(y, cols)
    values <- dataset[,y_pos]
    slices <- length(unique(values))
    sizes <- rep(0, slices)
    for(i in 1:slices) {
      sizes[i] <- table(values)[[i]]
    }
  }
  allocation <- c()
  for (i in 1:slices) {
    allocation <- allocation %>% append(rep(i, sizes[i]))
  }
  return(allocation)
}
```

## Means (used for the slicing)

```{r}
means <- function(dataset_one_slice) {
  nc <- ncol(dataset_one_slice)
  vals <- c()
  means_array <- colMeans(dataset_one_slice)
  for (i in 1:nc) {
    vals <- vals %>% append(means_array[[i]])
  }
  return (vals)
}

```

## Slicer

```{r}
sir_slicer_w2 <- function(dataset, categorical = FALSE, slices = 8, y, ordinal = FALSE) {
  dataset <- dataset %>% arrange(get(y))
  
  cols <- colnames(dataset)
  y_pos <- match(y, cols) # find index of column y
  
  if (categorical) {
    slices <- length(unique(dataset[,y_pos])) 
  }
  
  allocation <- divider_w2(dataset = dataset, slices = slices, categorical = categorical, y = y)
  dataset$slice <- allocation # add column to dataset giving which slice each observation goes to
  
  long_values_sliced_dataframe <- c()
  
  for (s in 1:slices) { 
    # find first row for current slice
    for (i in 1:nrow(dataset)) {
      if (dataset$slice[i] == s) {
        first_row <- i
        break
      }
    }
    # find last row for current slice
    for (i in 1:nrow(dataset)) {
      if (dataset$slice[i] == s) {
        last_row <- i
      }
    }
    # find average for all columns for the particular slice
    dataset_for_avging <- dataset[first_row:last_row, -ncol(dataset)][,-y_pos]
    avgs <- means(dataset_for_avging) # colMeans does same thing
    long_values_sliced_dataframe <- long_values_sliced_dataframe %>% append(avgs)
  }
  sliced_dataset <- matrix(long_values_sliced_dataframe, ncol = ncol(dataset)-2, byrow = TRUE) %>% as.data.frame()
  colnames(sliced_dataset) <- colnames(dataset)[-ncol(dataset)][-y_pos]
  rownames(sliced_dataset) <- c(1:slices)
  return(sliced_dataset)
}
```

## PCA

```{r}
sir_PCA <- function(dataset, directions) {
  all_pc <- prcomp(dataset)$rotation
  d <- min(ncol(all_pc), directions)
  return(all_pc[,1:d])
}
```

## Multiplier

```{r}
multiplier <- function(dataset, pc_dirs) {
  cov_mat <- cov(dataset)
  multiplied <- solve(cov_mat, pc_dirs)
  return(multiplied)
}
```

## Entire Function

```{r}
sir_all <- function(dataset, y = colnames(dataset)[1], slices = 8, directions = 10, categorical = FALSE) {
  
  sliced_dataset <- sir_slicer_w2(dataset = dataset, y = y, categorical = categorical, slices = slices)
  
  cols <- colnames(dataset)
  y_pos <- match(y, cols)
  
  pc_dirs <- sir_PCA(sliced_dataset, directions = directions)
  
  multiplied <- multiplier(data = dataset[,-y_pos], pc_dirs = pc_dirs)
  
  final <- as.matrix(dataset[,-y_pos]) %*% multiplied
  
  return(list(final, multiplied))
}
```

# Data Generation Functions

## Generator for X

```{r}
x_gen <- function(n, p) { # make sure it follows positive definite through mat %*% t(mat)
  m <- matrix(rnorm(p^2, 0, 5), ncol = p)
  sigmaX <- m %*% t(m)
  df <- rmvnorm(n, mean = rep(0, p), sigma = sigmaX) %>% as.data.frame()
  colnames(df) <- var_names(p)
  return(df)
}
```

## Generator for Y

```{r}
true_B <- c(1,0,0,3,1)/50
y_gen <- function(dataset, true_b = true_B) {
  nc <- ncol(dataset)
  nr <- nrow(dataset)
  y_vals <- rep(0, nr)
  
  y_vals <- sin(as.matrix(dataset) %*% true_b) + 0.01*rnorm(nr)
  
  dataset$y_vals <- y_vals
  return(dataset)
}

# comparable methods when max var y corred with var x 

# Compare ordinal with categorical (that has no order) - see how different in algorithm

# which ordinal categories to combine if need to combine some 
```

# Spatial .R file functions 

```{r}
x_sorter <- function(X, coords) {
  x <- coords[,1]
  X$x <- x
  X_sorted <- X %>% arrange(x)
  X <- X_sorted[,-ncol(X_sorted)]
  return(X)
}

y_sorter <- function(X, coords) {
  y <- coords[,2]
  X$y <- y
  X_sorted <- X %>% arrange(y)
  X <- X_sorted[,-ncol(X_sorted)]
  return(X)
}

spatial_allocator <- function(X, coords, slices = 8) {
  X <- x_sorter(X, coords)
  x_allocation <- allocator(X = X, slices = slices)
  X$x_slice <- x_allocation
  
  X <- y_sorter(X, coords)
  y_allocation <- allocator(X = X, slices = slices)
  X$y_slice <- y_allocation
  
  X$coordinate <- paste0(X$x_slice, ", ", X$y_slice)
  
  X <- X[,-c((ncol(X)-2),(ncol(X)-1))]
  return(X)
}

spatial_slicer <- function(X, coords, slices = 8) {
  X_allocated <- spatial_allocator(X = X, coords = coords, slices = slices)
  X_slice_sorted <- arrange(X_allocated, coordinate)
  slice_names <- unique(X_allocated$coordinate)
  long_avgs <- c()
  for (i in slice_names) {
    for (j in 1:nrow(X)) {
      if (X_slice_sorted$coordinate[j] == i) {
        first_row = j
        break
      }
    }
    for (j in 1:nrow(X)) {
      if (X_slice_sorted$coordinate[j] == i) {
        last_row = j
      }
    }
    dataset_for_avging <- X_slice_sorted[first_row:last_row,-ncol(X_slice_sorted)]
    avgs <- setNames(colMeans(dataset_for_avging), NULL) # colMeans does same thing
    long_avgs <- long_avgs %>% append(avgs)
  }
  sliced_dataset <- matrix(long_avgs, ncol = ncol(X), nrow = length(slice_names), byrow = TRUE) %>% as.data.frame()
  #rownames(sliced_dataset) <- c(1:slices)
  return(sliced_dataset)
}

spatial_sir <- function(X, coords, slices = 8, directions = 10, W = diag(slices^2)) {
  sliced_data <- spatial_slicer(X = X, coords = coords, slices = slices)
  pc_dirs <- sir_PCA(sliced_data, directions = directions, W = W)
  multiplied <- multiplier(data = X, pc_dirs = pc_dirs)
  final_XB <- as.matrix(X) %*% multiplied
  return(list(final_XB, multiplied))
}
```

# Testing

Testing with simple dataframes to check that the functions work and see basic concepts.

# Testing Spatial

```{r}
X_test <- x_gen(100, 5)
coords_test <- matrix(NA, nrow = 100, ncol = 2) %>% as.data.frame()
colnames(coords_test) <- c("x", "y")
coords_test$x <- sin(X_test$var1 + 3*X_test$var2 - 4*X_test$var4)
coords_test$y <- sin(X_test$var2 + 3*X_test$var3 - 4*X_test$var5)
rownames(X_test) <- c(1:100)
```


```{r}
sliced_spatial <- spatial_slicer(X = X_test, coords = coords_test, slices = 5)

sir_spatial <- spatial_sir(X = X_test, coords = coords_test, slices = 5, directions = 8)
```

## Var name function

Useful for making simple dataframes for testing.

```{r}
var_names <- function(nc) {
  vars <- c()
  for (i in 1:nc) {
    vars <- vars %>% append(paste0("var", i))
  }
  return(vars)
}
```

## dataframe_test

```{r}
dataframe_test <- as.data.frame(matrix(rnorm(20, 0, 5), nrow = 4))
nc <- ncol(dataframe_test)
vars <- c()
for(i in 1:nc) {
  vars <- vars %>% append(paste0("var", i))
}
colnames(dataframe_test) <- var_names(5)
```

## categorical_df

```{r}
categorical_df <- x_gen(6,5) %>% as.data.frame()
categorical_df <- categorical_df %>% cbind(c("apple", "apple", "orange", "orange", "apple", "orange"))
colnames(categorical_df) <- var_names(6)

sir_all(categorical_df, slices = 2, directions = 2, y = "var6")
```

# Case study: comparing SIR with PCA with trigonometric functions

Below are three functions with which we generate Y from a randomly generated X (following positive definite and other requirements). 

For all of the functions, I plotted the true Y values against the first dimension reduction direction. It does well if the relationship between the x and y axes on the plots is the same as the given function for the section (e.g if there appears to be a sine wave rather than random noise).

## Compare Y = sin(X)

```{r}
test_data <- x_gen(200, 5)
test_data <- test_data %>% y_gen(true_b <- c(1,0,0,3,1)/30)
colnames(test_data)[6] <- c("y_vals")

#head(sir_all(test_data, y = "y_vals", directions = 3, slices = 10)) # can run this to check it works 

sin_sir <- univariate_sir(X = test_data[,-6], Y = as.data.frame(test_data[,6]), slices = 10, directions = 3)[[1]]

length(sin_sir[,1])
length(test_data$y_vals)
length(pr[,1])

sin_df <- matrix(NA, nrow = 200, ncol = 3) %>% as.data.frame()
colnames(sin_df) <- c("y", "pca", "sir")
sin_df$y <- test_data$y_vals
sin_df$sir <- sin_sir[,1]
sin_df$pca <- pr[,1]

ggplot(sin_df, aes(x = sir, y = y)) + 
  geom_point(col = "#e64626") +
  labs(x = "First SIR Direction", y = "True Y Values") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 17),
        axis.title.y = element_text(size = 17),
        axis.title.x = element_text(size = 17))
  
ggplot(sin_df, aes(x = pca, y = y)) + 
  geom_point(col = "#e64626") +
  labs(x = "First Principal Component", y = "True Y Values") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 17),
        axis.title.y = element_text(size = 17),
        axis.title.x = element_text(size = 17))

plot(sin_sir[,1], test_data$y_vals, main = "Y and first SIR direction", ylab = "True Y values", xlab = "X projected along first SIR direction") # plot 1: sine wave 

pr <- prcomp(test_data[,-6])$x

plot(pr[,1], test_data$y_vals, main = "Y and first PC", xlab = "First Principal Component", ylab = "True Y values") # plot 2. Compare to plot 1, does not look like the pr[,1] vector captures the information of X well at all. If it did, this would look like a sine wave.

dataframe_check <- matrix(c(scale(multipd2[,1]), scale(xb[,1]), test_data$y_vals), ncol = 3, byrow = FALSE) %>% as.data.frame()
colnames(dataframe_check) <- c("mine", "package", "y_vals")

ggplot(dataframe_check) +
  geom_point(data = dataframe_check, aes(x = mine, y = y_vals, color = "red")) +
  geom_point(data = dataframe_check, aes(x = package, y = y_vals, color = "blue"))
scaled_dr_obj <- multipd3$evectors[,1] %>% setNames(NULL) %>% scale()
#scaled_dr_obj[1:5] %>% plot()
scaled_my_obj <- sir_all_obj[2][[1]][,1] %>% setNames(NULL) %>% scale()
scaled_my_obj[1:5] %>% plot(col = "red")
points(scaled_dr_obj[1:5], col = "blue")
```

Why is a sine wave above good? If we could theoretically perfectly reduce the dimension of X (n * p matrix) to a n * 1 vector named x, we would get a sine wave since Y = sin(x). What we have done is use SIR to reduce X to a n * 1 vector that is PC1, and by plotting PC1 against Y we have produced a graph that looks a lot like Y = sin(PC1). Therefore, it looks like the dimension reduction technique performed well. 

```{r}
cor(true_score, multipd[,1])
cor(true_score, pr[,1])
```

## Y = cos(X)

```{r}
set.seed(2023)
y_gen_cos <- function(dataset) {
  nc <- ncol(dataset)
  nr <- nrow(dataset)
  y_vals <- rep(0, nr)
  for (i in 1:nr) {
    y_vals[i] <- rnorm(1, 0, 0.01) + sin((dataset[i, 1] + 3*dataset[i, 4] + dataset[i, 5])/1000 + 0.45) #true score comes from here - this direction times x
  }
  dataset$y_vals <- y_vals
  return(dataset)
}
true_B <- c(1,0,0,3,1)/50

test_data_cos <- x_gen(200, 5) %>% y_gen_cos()

true_score <- as.matrix(test_data_cos[,-6]) %*% true_B

data_cos_SIRed <- sir_all(test_data_cos, directions = 3, slices = 10, categorical = FALSE, y = "y_vals")

cor(data_cos_SIRed[,1], true_score)

pc_cos <- prcomp(test_data_cos[,-6])$x

cor(pc_cos[,1], true_score)

plot(data_cos_SIRed[,1], test_data_cos$y_vals)

plot(pc_cos[,1], test_data_cos$y_vals)
```

## Y Binary

Not binary, but discretises continuous x values into y = c(1,2,3,4). Performs the same task as binary but better.

The better performance here can be seen in the fact that we have less overlap between the x values of the y categories. In the true function, the lowest x values are given y = 0, the next x values are given y = 1, and so on. The SIR DR preserves that relationship better than the PCA DR.

```{r}
y_bin <- function(dataset) {
  nc <- ncol(dataset)
  nr <- nrow(dataset)
  y_vals <- rep(0, nr)
  for (i in 1:nr) {
    value <- rnorm(1, 0, 0.001) + sin((dataset[i, 1] + 3*dataset[i, 4] + dataset[i, 5])/50)
    if (value > 0.4) {
      y_vals[i] <- 3
    } else if ((value <= 0.4) & (value > 0)) {
      y_vals[i] <- 2
    } else if ((value <= 0) & (value > -0.4)) {
      y_vals[i] <- 1
    } else {
      y_vals[i] <- 0
    }
  }
  dataset$y_vals <- factor(y_vals)
  return(dataset)
}

# discrete formula: 
# 1) randomly generate X
# 2) create value = sin(XB)
# 3) 

bin_dataset <- x_gen(n = 1000, p = 5) %>% y_bin
bin_dataset

bin_x <- bin_dataset %>% dplyr::select(-c(y_vals))
bin_y <- bin_dataset %>% dplyr::select(c(y_vals))
bin_x
bin_y
bin_sired <- univariate_sir(bin_x, Y = bin_y, slices = 10, directions = 3)

plot(bin_sired[[1]][,1], bin_dataset$y_vals)

pc <- prcomp(bin_dataset[,-6])$x
plot(pc[,1], bin_dataset$y_vals)

discrete_for_plot <- matrix(NA, nrow = 2000, ncol = 3) %>% as.data.frame()
colnames(discrete_for_plot) <- c("y_vals", "method", "z_val")
discrete_for_plot$y_vals <- rep(bin_dataset$y_vals, 2)
discrete_for_plot$method <- c(rep("PCA", 1000), rep("SIR", 1000))
discrete_for_plot$z_val <- c(pc[,1], bin_sired[[1]][,1]*100)
ggplot(discrete_for_plot, aes(x = z_val, y = y_vals)) +
  geom_point(col = "#e64626") +
  facet_wrap(~method, scales = "free") +
  labs(x = "First Direction of Z", y = "True Y Values") +
  ggtitle("First Direction of Z from PCA, SIR DR against True Y Values") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
  

bin_dr_onj <- dr(y_vals ~ var1 + var2 + var3 + var4 + var5, data = bin_dataset, method = "sir", numdir = 5, nslices = 4)
bin_dr_onj$slice.info

Y_bin <- as.data.frame(bin_dataset[,6])
X_bin <- bin_dataset[,1:5]
my_sir_bin_obj <- categorical_sir(X = X_bin, Y = Y_bin, directions = 5)


plot(scale(bin_dr_onj$evectors[,1])[1:5], col = "red")
points(scale(my_sir_bin_obj[[2]][,1])[1:5], col = "blue")

bin_dr_onj$evectors[,1]/sqrt(sum(bin_dr_onj$evectors[,1]^2))
my_sir_bin_obj[[2]][,1]/sqrt(sum(my_sir_bin_obj[[2]][,1]^2))
```

# Comparison between .R function and dr package

```{r}
dr_obj <- dr(formula = y~var1+var2+var3+var4+var5, data = test_data, method = "sir", nslices = 10, numdir = 4) %>% summary
dr_obj
sir_obj <- univariate_sir(X = test_data[,-6], Y = as.data.frame(test_data[,6]), directions = 4, slices = 10)[[1]]

plot(sir_obj[,1], test_data$y)
```





























